A Plasma Torus Around a Young Low-Mass Star

### 2025-04-22 paper/ms.tex
update

\documentclass{nature3}
\usepackage{graphicx}
\usepackage{float}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{aas_macros_nature}
\usepackage{lineno}
\usepackage{caption}
\usepackage{rotating} % sidewaystable

@@

\newcommand{\farcm}{\mbox{\ensuremath{.\mkern-4mu^\prime}}}%    % fractional arcminute symbol: 0.'0
\newcommand{\farcs}{\mbox{\ensuremath{.\!\!^{\prime\prime}}}}%  % fractional arcsecond symbol: 0.''0

@@

@@

\author{Luke G. Bouma$^{1,2}$}

\maketitle

\scriptsize
\begin{affiliations}
\item \carnegie
\item \caltech
\item Carnegie Fellow; 51 Pegasi b Fellow
\end{affiliations}
\normalsize

@@

\begin{abstract}
\normalfont
A small fraction of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, which provides the first direct evidence for
corotating plasma clumps around a CPV.  While cool ($10^4$ K)
plasma can persist within the hot ($10^6$ K) coronae of
rapidly-rotating stars across a broad range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
our observations suggest that red dwarfs in particular can
sustain long-lived, optically thick clumps that generate the 
sharp photometric features seen in CPVs.  The origin of this
optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

\maketitle

@@

% Main text – up to 3,000 words, excluding abstract, Methods,
% references and figure legends.

\section{Main}
\label{sec:main}

M dwarfs, stars with masses below about half that of the Sun, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water \cite{NAP26141}.
Community investment with JWST has proceeded accordingly
\cite{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  How an M
dwarf's evolution influences its planets---especially the retention of
their atmospheres--- has in turn become a major theme in exoplanet and
stellar astrophysics.  Previous work has established that most M
dwarfs host close-in planets \cite{Dressing2015} that on average are
subject to long circumstellar disk lifetimes \cite{Ribas2015}, to high
doses of UV radiation \cite{France2016}, and to a high incidence of
flares and coronal mass ejections \cite{Gunther2020}.  However,
despite extensive work in these areas, the plasma and magnetospheric
environments that bathe young, close-in exoplanets remain challenging
to quantify.  Understanding these environments is crucial because they
directly impact atmospheric retention and habitability of close-in
planets.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  While Figure~\ref{fig:lc} highlights the main
object of interest in this article, over one hundred analogous objects
have now been found by K2 and TESS
\cite{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Bouma2024}.
These CPVs are phenomenologically identified based on their
structured, periodic optical light curves; most are M dwarfs with
rotation periods shorter than two days.  Within current sensitivity
limits, none host disks \cite{Stauffer2017,Bouma2024}.  However,
$\approx$3\% of stars a few million years old show this complex
photometric behavior, an observed fraction which decreases to
$\approx$0.3\% by $\approx$150\,Myr \cite{Rebull2020}.  CPVs are a
known source of astrophysical false positives in searches for
transiting exoplanets around young M dwarfs
\cite{vanEyken2012,Johns-Krull2016,Bouma2020}.

The two leading hypotheses for explaining CPVs are either that
transiting clumps of circumstellar material corotate with the star
\cite{Stauffer2017,Gunther2022,Bouma2024}, or that these stars
represent an extreme in naturally-occurring distributions of starspots
or faculae \cite{Koen2021}.  The main argument against a starspot-only
explanation invokes the timescales and amplitudes of the sharpest
photometric features.  However, no independent evidence has yet been
acquired for the presence of any circumstellar material.
Geometrically, transiting clumps would imply an occurrence a few to at
most ten times the observed rate.  The clumps could therefore exist
around 10-30\% of M dwarfs during their early lives.

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/f1.pdf}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 1 (Movie):  TIC~141146667 is a complex periodic
  variable (CPV).}  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_20250116.mp4}{available here},
  covers a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period
  in groups of three cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute sampling
  are in gray; black averages to 100 points per cycle.  Similar to
  other stars of this class, the sharp photometric features persist
  for tens to thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy of
stars with a wide range of masses has shown periodic high-velocity
absorption and emission in Balmer lines such as H$\alpha$, interpreted
as long-lived, corotating clumps of cool plasma
\cite{CollierCameron1989,CollierCameron1992,Barnes2000,Donati2000,Dunstone2006,Skelly2008,Leitzinger2016,Cang2021}.
Such clumps are thought to be forced into corotation by the magnetic
field, and the exact geometry of where the plasma can accumulate is
dictated by the field's topology.  For instance, a magnetic dipole
field tilted with respect to the stellar spin axis yields
accumulations in a warped torus geometry \cite{Townsend2005}, whereas
in the limit of a single strong field line, accumulation occurs at the
line's apex, furthest from the star \cite{Waugh2022}.  To date, none
of these spectroscopic variables have shown any photometric anomalies
\cite{Bouma2024}, leaving open the issue of whether they are related
to CPVs.

In this study, we present the first spectroscopic detection of
corotating clumps of cool plasma around a CPV.  We identified
TIC~141146667 in previous work \cite{Bouma2024} by searching the TESS
two-minute data \cite{Ricker2015} for stars showing highly structured,
periodic light curves.  We selected the star for spectroscopy because
its brightness and rotation rate enabled an efficient search for
variability in its line profiles.  We observed the star for five hours
on 17~February~2024 (UT) using the High Resolution Echelle
Spectrometer (HIRES \cite{vogt_hires_1994}) on the 10\,m Keck I
telescope.  TESS observed the star over a four week interval from
30~January~2024 to 26~February~2024 with a duty cycle of 77\%.  During
the HIRES observations, scattered light from the Earth caused a gap in
the TESS photometry; TESS data collection resumed 12 hours (three
rotation cycles) after the spectra were acquired.  Extended Data
Figure~\ref{fig:fulllc} shows the detailed photometric behavior of the
star near the epoch of observation; the general photometric morphology
is similar before and after the data gap.

\section{Results}

\begin{figure}[!tp]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/f2.pdf}
  \captionsetup{labelformat=empty}
  \caption{{\bf Figure 2 (Movie):}  
  Emission from circumstellar plasma orbiting TIC 141146667.
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{available here},
  shows the spectral evolution over five hours.
  {\bf a,} Average TESS light curve from 5 February 2024 to 26
  February 2024 folded on the \periodhr\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf b,} Keck/HIRES H$\alpha$ spectra acquired on 17
  February 2024.  The continuum is set to unity, and the darkest color
  is at twice the continuum to accentuate emission outside the line
  core ($|v/v_{\rm eq}|>1$, for $v_{\rm eq}$=130\,\kms).  While
  emission in the line core originates in the star's chromosphere,
  the sinusoidal emission features are most readily described by a
  warped plasma torus.
  {\bf c,} Individual epochs of Panel b, visible in the
  online movie.  The dotted line shows the time-averaged spectrum,
  $f_{\langle t \rangle}$.
  {\bf d,} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf e,} As in Panel b, after subtracting the time-averaged
  spectrum.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch
  is set to mirror the dynamic range of the data.
  {\bf f,} Individual epochs of Panel e, visible in the online
  movie.}
  \label{fig:spec}
\end{figure}

% photometry
Figure~\ref{fig:spec} compiles the TESS and HIRES data from February
2024.  Over timescales of years, CPVs maintain fixed periods while
their photometric morphology evolves.  TIC~141146667 indeed evolved
relative to the February 2022 discovery data (see Figure~\ref{fig:lc}).
In February 2024, the average photometric signal showed a small
brightening over 45\% of the period, followed by a complex flux dip
spanning 55\% of the period.  This eclipse feature shows two to three
local photometric minima, and one to two local maxima.  Its W-shape is
suggestive of eclipse geometries seen in forward models of warped
plasma tori (see \cite{Townsend2008} and the associated
\href{http://user.astro.wisc.edu/~townsend/static.php?ref=rrm-movies#Download_Bundles}{movies}).
% particularly ω/ωc = 0.5, beta = 30, i = 90.

The spectroscopy shows emission beyond the star's equatorial velocity
($v_{\rm eq}$=130\,\kms).  There are at least two distinct emission
components, separated by half a cycle in phase.  The inner component
at lower velocities has clearer sinusoidal behaviour in time and is
double-peaked, with peak semi-amplitudes of 2.07\,$v_{\rm eq}$ and
2.88\,$v_{\rm eq}$ (see \S~\ref{subsec:halpha}).  There is significant
non-periodic variability in the emissivity of this double-peaked
component: the flux excess begins with an amplitude 70\% that of the
continuum early in the observation sequence, and diminishes to 10\% by
its end.  The higher-velocity component 180$^\circ$ opposite in phase
is detected only from $\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this outer
component appears connected to the star in velocity space.  While its
peak semi-amplitude of 3.88\,$v_{\rm eq}$ is achieved at both
$\phi$=0.25 and 0.75, its amplitude similarly decreases from a 60\%
excess over the continuum at the beginning of the observation sequence
to a 10\% excess by its end.  The apparent period for all three
spectroscopic emission components is consistent with the photometric
\periodhr\ hour period.

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
observed sinusoidal periods and velocities, this material's motion is
not controlled by gravitational attraction to the star; it is more
readily explained as a partially-ionized plasma being dragged along
with a rigidly rotating stellar magnetic field.  The velocity
semi-amplitude of the sinusoids gives the mean distance of each clump
from the stellar surface: 2.07\,$R_\star$ and 2.88\,$R_\star$ for the
inner clumps, and 3.88\,$R_\star$ for the outer clump.   These clumps
transit in front of the star when passing from negative to positive
velocity.  The transits of the two inner clumps last $\approx$22\% of
each cycle, from $\phi$=-0.1 to $\phi$=+0.12 (see
\S~\ref{subsec:halpha}).  This spectroscopic transit coincides with
the latter half of the complex eclipse feature in the TESS data.

@@

\begin{table*}

@@

\caption{Selected system parameters for TIC~141146667}

@@

1: SED fit \cite{Bouma2024}.

@@

   while the spectrum lacks lithium (see \S~\ref{subsec:stparams}).
3: PARSEC v1.2S \cite{Chen2014}.

@@

7: Gaia DR3 geometric \cite{GaiaDR3}.

@@

\end{table*}

\section{Discussion}

Spectra of magnetically-active, rapidly rotating stars with a wide
range of masses have been previously observed to exhibit both
sinusoidal time-varying Balmer emission
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and transient
absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020}, similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for this
spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar
corona that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's photosphere because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate in a
centrifugally-supported magnetosphere
\cite{Petit2013,Daley-Yates2024}.  Generally speaking, such material
need neither transit nor be opaque in broadband optical light.  In
fact, Figure~\ref{fig:spec} suggests that the transits of the
H$\alpha$ emitting material cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast,
and at orbital phases that only partially overlap with the photometric
modulations.

Nonetheless, our Keck/HIRES observations are the first reported
time-series spectra of a CPV, and they demonstrate that
corotating circumstellar plasma clumps exist around at least one such
star.  While starspots likely do contribute smooth signals to the
photometric variability, a ``starspot-only'' scenario for the CPVs
\cite{Koen2021} is incompatible with our observations.
Scenarios in which the circumstellar material is made only of dust are
similarly ruled out.  While starspots and dust may both be present,
the observed H$\alpha$ emission requires clumps of plasma with a
significant population of hydrogen atoms in the $n$=3 excited state,
with most H$\alpha$ emission in each clump coming from a region half
the size of the star (see \S~\ref{subsec:halpha}).  The observed
H$\alpha$ luminosity suggests characteristic densities and masses for
the gaseous component of these clumps of $n_{\rm H} \sim
10^{11}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
\S~\ref{subsec:gas}).  Dust, if present, is independently constrained
to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm
g}$ (see \S~\ref{subsec:dust}).  This circumstellar material could
originate either from the star or from an external source.  Plausible
external sources include an undetected old disk, comets, or a close-in
exoplanet.  This latter scenario would make CPVs extrasolar analogs of
the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

>>>>

% upon AAS submission
\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex7}
%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex7}
% drafting / arxiv
%\documentclass[11pt,twocolumn,modern]{aastex7}
\turnoffedit

\usepackage{apjfonts}
\usepackage{url}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath,amstext,amssymb}
\usepackage[caption=false]{subfig} % for subfloat
\usepackage{xcolor, fontawesome}
\usepackage{color}
\usepackage{enumitem}
\usepackage{rotating} % sidewaystable

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\correspondingauthor{Luke G. Bouma}

\received{--}
\revised{--}
\accepted{--}
\shorttitle{M Dwarf Plasma Torus}

\shortauthors{Bouma et al.}

\author[orcid=0000-0002-0514-5538,sname='Bouma']{Luke~G.~Bouma}
\altaffiliation{Carnegie Fellow; 51 Pegasi b Fellow}
\affiliation{\carnegie}
\affiliation{\caltech}
\email{lbouma@carnegiescience.edu}

@@

\begin{abstract}
A small fraction of red dwarfs younger than $\sim$100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material that corotates with the star.  However,
the composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces.  Here, we present
time-series spectroscopy and photometry of a rapidly-rotating
($P$=3.9\,hr) CPV, TIC\,141146667.  The spectra show sinusoidal
time-varying Balmer emission at twice to four times the star's
equatorial velocity, providing direct evidence for cool
($\lesssim$10$^4$\,K) plasma clumps trapped in corotation around a CPV.
These data support the idea that young, rapidly-rotating M dwarfs can
sustain warped plasma tori, similar to massive magnetic stars.  The
origin of the optically thick material in these tori, and the issue
of whether this material contains a significant amount of dust,
remain unclear.
\end{abstract}

\keywords{Circumstellar matter (241), Stellar magnetic fields (1610),
Stellar rotation (1629) Periodic variable stars (1213), Weak-line T
Tauri stars (1795) }

@@

\section{Introduction}
\label{sec:intro}

Stars with masses below about half that of the Sun, M dwarfs, are the
only type of star to offer near-term prospects for detecting the
atmospheres of rocky exoplanets with surface water.  Community
investment with JWST is proceeding accordingly
\citep[see][]{Redfield2024,TRAPPIST1JWSTCommunityInitiative2024}.  The
question of how an M dwarf's evolution influences its
planets---especially the retention of their atmospheres---has in turn
become a major theme in exoplanet and stellar astrophysics.  Previous
work has established that most M dwarfs host close-in planets
\citep{Dressing2015} that on average are subject to long circumstellar
disk lifetimes \citep{Ribas2015}, to high doses of UV radiation
\citep{France2016}, and to a high incidence of flares and coronal mass
ejections \citep{Feinstein2020}.  However, despite extensive work in
these areas, the plasma and magnetospheric environments that bathe
young, close-in exoplanets remain challenging to quantify.
Understanding these environments is crucial because they directly
impact atmospheric retention and habitability of close-in planets.

One glaring example of our current ignorance is the complex periodic
variables (CPVs).  While Figure~\ref{fig:lc} highlights the main
object of interest in this article, over one hundred analogous objects
have now been found by K2 and TESS
\citep{Rebull2016,Stauffer2017,Rebull2018,Zhan2019,Rebull2020,Stauffer2021,Popinchalk2023,Bouma2024}.
These CPVs are phenomenologically identified based on their
structured, periodic optical light curves; most are M dwarfs with
rotation periods shorter than two days.  Within current sensitivity
limits, none host disks \citep{Stauffer2017,Bouma2024}.  However,
$\approx$3\% of stars a few million years old show this complex
photometric behavior, an observed fraction which decreases to
$\approx$0.3\% by $\approx$150\,Myr \citep{Rebull2020}.  CPVs are a
known source of astrophysical false positives in searches for
transiting exoplanets around young M dwarfs
\citep{vanEyken2012,Johns-Krull2016,Bouma2020}.

The two leading hypotheses for explaining CPVs are either that
transiting clumps of circumstellar material corotate with the star
\citep{Stauffer2017,Gunther2022,Bouma2024}, or that these stars
represent an extreme in naturally-occurring distributions of starspots
or faculae \citep{Koen2021}.  The main argument against a
starspot-only explanation invokes the timescales and amplitudes of the
sharpest photometric features.  However, no independent evidence has
yet been acquired for the presence of any circumstellar material.  If
the material exists, then the geometric correction from the transit
probability would imply an intrinsic occurrence rate at least a few
times larger than the observed rate, suggesting that these clumps
could exist around $\gtrsim$10\% of M dwarfs during their early lives.

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young stars have, for
decades, reported that stellar coronae contain both hot
($\gtrsim$$10^6$ K) and cool ($\lesssim$$10^4$ K) plasma. In
particular, time-series spectroscopy of stars with a wide range of
masses has shown periodic high-velocity absorption and emission in
Balmer lines such as H$\alpha$, interpreted as long-lived, corotating
clumps of cool plasma
\citep{CollierCameron1989,CollierCameron1992,Barnes2000,Donati2000,Dunstone2006,Skelly2008,Leitzinger2016,Cang2021}.
Such clumps are thought to be forced into corotation by the magnetic
field, and the exact geometry of where the plasma can accumulate is
dictated by the field's topology.  For instance, a magnetic dipole
field tilted with respect to the stellar spin axis yields
accumulations in a warped torus geometry \citep{Townsend2005}, whereas
in the limit of a single strong field line, accumulation occurs at the
line's apex, furthest from the star \citep{Waugh2022}.  To date, none
of these spectroscopic variables have shown any photometric anomalies
\citep{Bouma2024}, leaving open the issue of whether they are related
to CPVs.

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.47\textwidth]{figures/f1.pdf}
  \vspace{-0.3cm}
  \caption{{\bf Figure 1 (Movie):  TIC\,141146667 is a complex periodic
  variable (CPV).}  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_20250116.mp4}{available here},
  covers a baseline of 5{,}784 cycles irregularly sampled over three
  years.  The TESS light curve is phased to the \periodhr\ hour period
  in groups of three cycles per frame.  This is the period both of
  stellar rotation, and (we hypothesize) of corotating clumps of
  circumstellar material.  Raw data acquired with two minute sampling
  are in gray; black averages to 100 points per cycle.  Similar to
  other stars of this class, the sharp photometric features persist
  for tens to thousands of rotational cycles. }
  \label{fig:lc}
\end{figure}

In this study, we present the first spectroscopic detection of
corotating clumps of cool plasma around a CPV, TIC\,141146667.  The
observations are presented in Section~\ref{sec:obs}, with the main new
evidence in Figure~\ref{fig:spec}.  Section~\ref{sec:results} expounds
on the results; Section~\ref{sec:disc} discusses their interpretation
and highlights future directions.

\section{Observations}
\label{sec:obs}

We identified TIC\,141146667 in previous work \citep{Bouma2024} by
searching TESS two-minute data from 2018-2022 for highly
structured, periodic light curves \citep{Ricker2015}.  We selected the
star for spectroscopy because its brightness and rotation rate enabled
an efficient search for variability in its line profiles.
As apparently single pre-main sequence M dwarf, its properties are
typical for the general CPV population (see Table~\ref{tab:params} and
Appendix~\ref{sec:stparams}).

We observed TIC\,141146667 ($V$=16.2) for five hours on
17~February~2024 (UT) using the High Resolution Echelle Spectrometer
(HIRES; \citealt{vogt_hires_1994}) on the 10\,m Keck I telescope.  The
observations spanned the second half of the night, from 17 February
2024 10:47 to 16:13 (UT).  The star's airmass spanned $z$=1.2-2.2, and
we opted for a fixed 15 minute cadence, except for a final 10 minute
exposure due to increasing sky brightness at sunrise.  We observed
without the iodine cell and used the C2 decker
(0$\farcs$86$\times$14$\farcs$0) in the red configuration, yielding a
spectral resolution $R$$\approx$45{,}000 ($\delta
v$$\approx$6.7\,\kms; $\delta v / v_{\rm eq}$$\approx$0.05).  We
binned the CCD readout by a factor of three in the spatial dimension,
yielding $\approx$1,000 photons (S/N=33) per pixel in the continuum at
6500\,\AA, at minimum airmass.  Strong winds contributed to
1$''$-1\farcs5 seeing over the night, but conditions were otherwise
favorable.  We reduced the echelleogram to a one-dimensional spectrum
using the standard techniques of the California Planet Survey
\citep{Howard2010}.  Figure~\ref{fig:spec} shows the result in the
vicinity of H$\alpha$ without any additional processing.

To date, TESS has observed TIC\,141146667 ($T$=13.3) for a total of six
non-contiguous months spanning 2019-2024.  Two-minute cadence observations were
acquired during Sectors 41, 48 (TESS DDT039, PI: Kunimoto), and 75
(TESS Program G06030, PI: Bouma); the remaining 30-minute data during
Sectors 14, 15, and 21 smears sharp features over the $<$4\,hour
period (see \citealt{Gunther2022}).  The concatenated two-minute data
are shown in the movie associated with Figure~\ref{fig:lc}.  The movie shows that,
like other CPVs, TIC\,141146667 maintains a fixed period over
timescales of years while its detailed photometric morphology evolves.
The nearest known star, TIC\,141146666 ($T$=14.5), is 25$''$ from
TIC\,141146667 and is photometrically stable in the TESS target pixel
cutouts; crowding is not a concern.

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.95\textwidth]{figures/sf1.pdf}
  \vspace{-0.2cm}
  \caption{{\bf Evolution of TIC\,141146667 during the Keck/HIRES
  observation (green bar).}  {\bf a,} TESS simple aperture photometry.
  Data gaps caused by light from the Earth (BTJD 3356-3358.5) and Moon
  (BTJD 3359.5-3360.5) are visible.  Raw two minute
  data are in gray; black time-averages to ten minute sampling.  {\bf
  b-c,} Folded TESS light curve before and after spectroscopy.  Black
  now phase-averages to 100 points per
  cycle.  During BTJD 3352-3356, a state switch occurred near BTJD
  3353, and the dip at $\phi$$\approx$0.8 disappeared.  While the
  large flux decrement was present both before and after the HIRES
  sequence, its photometric shape evolved during the data gap.  The
  orange bar denotes times of spectroscopic transits for the inner two
  H$\alpha$ clumps observed with HIRES (see Figure~\ref{fig:spec}).}
  \label{fig:fulllc}
\end{figure*}

We observed with Keck during Sector~75 in an attempt to obtain
simultaneous observations.   Figure~\ref{fig:fulllc} shows the result;
the Earth passed within 25$^\circ$ of the
TESS camera's boresight from BTJD 3356.0~-~3358.5, which caused a data
gap that ended twelve hours after the Keck/HIRES observation epoch
(green bar).  From
BTJD 3359.4~-~3362.0, the Moon then passed within 25$^\circ$ of the
camera's boresight.  Based on the level of scattered light in the
optimal TIC\,141146667 aperture \citep{Jenkins2016}, we manually masked
out times from 3359.40~-~3360.13; the remainder of the data during the
lunar approach were usable.  Small data gaps from BTJD
3353.55~-~3353.77 and from BTJD 3360.12~-~3360.33 were caused by data
downlinks at the spacecraft's perigee and apogee, respectively.

While the data gap is unfortunate, Figure~\ref{fig:fulllc} shows
that a large, complex flux decrement spanned roughly half of each
cycle both before and after the HIRES data were acquired.  From BTJD
3352~-~3356, this dip had two sharp local minima;  the local minimum
at $\phi$$\approx$0.8 disappeared following the flare at BTJD 3353,
yielding a dip more closely resembling an asymmetric ``V'' than a ``W''.
Similar CPV state changes have been previously noted 
\citep{Stauffer2017,Bouma2024}.  The photometric shape therefore
evolved during the twelve cycles spanning the 3356~-~3358 gap, since
the average shape from 3358-3363 more closely resembles the initial
``W''.  Nonetheless, the general photometric morphology---a small
brightening over 45\% of the period, followed by a complex flux dip
spanning 55\% of the period---is similar before and after the data
gap.

\section{Results}
\label{sec:results}

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.99\textwidth]{figures/f2.pdf}
  \vspace{-0.2cm}
  \caption{{\bf Figure 2 (Movie):}  
  Emission from circumstellar plasma orbiting TIC\,141146667.
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{available here},
  shows the spectral evolution over five hours.
  {\bf a,} Average TESS light curve from 5 February 2024 to 26
  February 2024 folded on the \periodhr\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf b,} Keck/HIRES H$\alpha$ spectra acquired on 17
  February 2024.  The continuum is set to unity, and the darkest color
  is at twice the continuum to accentuate emission outside the line
  core ($|v/v_{\rm eq}|>1$, for $v_{\rm eq}$=130\,\kms).  While
  emission in the line core originates in the star's chromosphere,
  the sinusoidal emission features are most readily described by a
  warped plasma torus.
  {\bf c,} Individual epochs of Panel b, visible in the
  online movie.  The dotted line shows the time-averaged spectrum,
  $f_{\langle t \rangle}$.
  {\bf d,} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf e,} As in Panel b, after subtracting the time-averaged
  spectrum.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch
  is set to mirror the dynamic range of the data.
  {\bf f,} Individual epochs of Panel e, visible in the online
  movie.}
  \label{fig:spec}
\end{figure*}

Figure~\ref{fig:spec} compiles the TESS and HIRES data from February
2024.  The HIRES spectra show emission in H$\alpha$ beyond the star's
equatorial velocity, $v_{\rm eq}$, of 130\,\kms.  There are at least
two distinct emission components, separated by half a cycle in phase.
The inner component at lower velocities has clearer sinusoidal
behaviour in time and is double-peaked, with peak semi-amplitudes of
2.07\,$v_{\rm eq}$ and 2.88\,$v_{\rm eq}$, measured following the
procedure in Appendix~\ref{subsec:halpha}.  There is significant
non-periodic variability in the emissivity of this double-peaked
component: the flux excess begins with an amplitude 70\% that of the
continuum, and diminishes to 10\% by the end.  The higher-velocity
component 180$^\circ$ opposite in phase is detected only from
$\phi$=0.2-1.0.  From $\phi$=0.2-0.5, this outer component appears
connected to the star in velocity space.  While its peak
semi-amplitude of 3.88\,$v_{\rm eq}$ is achieved at both $\phi$=0.25
and 0.75, its amplitude similarly decreases from a 60\% excess over
the continuum at the beginning of the observation sequence to a 10\%
excess by its end.  
From the same procedure in Appendix~\ref{subsec:halpha},
we measured
the period for all three spectroscopic
emission components to be consistent with the photometric \periodhr\ hour
period, to within two minutes for the lower-velocity component.

These sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to corotate with the star.  Based on the
observed sinusoidal periods and velocities, this material's motion is
not controlled by gravitational attraction to the star; it is more
readily explained as a plasma being dragged along
with a rigidly rotating stellar magnetic field.  The velocity
semi-amplitude of the sinusoids gives the mean distance of each clump
from the stellar surface: 2.07\,$R_\star$ and 2.88\,$R_\star$ for the
inner clumps, and 3.88\,$R_\star$ for the outer clump.   These clumps
transit in front of the star when passing from negative to positive
velocity.  The transits of the two inner clumps last $\approx$22\% of
each cycle, from $\phi$=-0.1 to $\phi$=+0.12 (see
Appendix~\ref{subsec:halpha}).  This spectroscopic transit coincides with
the latter half of the complex eclipse feature in the TESS data.

@@

\begin{table}

@@

\caption{Selected system parameters for TIC\,141146667}

@@

1: SED fit \citep{Bouma2024}.

@@

   while the spectrum lacks lithium (see \S~\ref{sec:stparams}).
3: PARSEC v1.2S \citep{Chen2014}.

@@

7: Gaia DR3 geometric \citep{GaiaDR3}.

@@

\end{table}

\section{Discussion}
\label{sec:disc}

\subsection{Corotating Plasma Clumps}

Our Keck/HIRES observations are the first reported time-series spectra
of a CPV, and they demonstrate that corotating circumstellar plasma
clumps exist around at least one such star.  While starspots likely do
contribute smooth signals to the photometric variability, a
``starspot-only'' scenario for the CPVs \citep{Koen2021} is
incompatible with our observations.  Scenarios in which the
circumstellar material is made only of dust are similarly ruled out.
While starspots and dust may both be present, the observed H$\alpha$
emission requires plasma with a significant population of hydrogen
atoms in the $n$=3 excited state.  Most of this H$\alpha$ emission
originates in ``clumps'' with size similar to that of the star; radial
``spokes'' or azimuthally-extended ``arcs'' for the emitting material are
ruled out by the observed H$\alpha$ velocity dispersion (see
\S~\ref{subsec:halpha}).  
The observed H$\alpha$ luminosity suggests characteristic densities
and masses for the gaseous component of these clumps of $n_{\rm H}
\sim 10^{11}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
\S~\ref{subsec:gas}).  Dust, if present, is independently constrained
to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm
g}$ (see \S~\ref{subsec:dust}).  
The spectra show minimal evidence for significant populations in
higher-order hydrogen states (see Appendix~\ref{subsec:specvar}).

\subsection{Plasma Tori Could Explain Complex Periodic Variables}

While the H$\alpha$ emission appears consistent with clumps, the
photometric width of the complex flux dips in TIC\,141146667
specifically, and CPVs more broadly \citep[Figure~4 of][]{Bouma2024},
raise the possibility that azimuthally-distributed material may be the
norm for these objects.  In other words, a torus may be a more
accurate picture than a clump.

\citet{Townsend2005} outline the physics of how such structures can
exist: for stars with magnetospheric radii $R_{\rm m}$ that exceed
their corotation radii $R_{\rm c}$, the effective potential
experienced by a charged particle can have a local minimum outside
$R_{\rm c}$, enabling material to accumulate in a
centrifugally-supported magnetosphere
\citep{Petit2013,Daley-Yates2024}.  Generally speaking, such material
need neither transit nor be opaque in broadband optical light.  In
fact, Figure~\ref{fig:spec} suggests that the transits of the
H$\alpha$ emitting material cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast,
and at orbital phases that only partially overlap with the photometric
modulations.

Nonetheless, \citet{Townsend2008} synthesized light curves for plasma
trapped in rigidly rotating magnetospheres of rapidly rotating stars;
and found that assuming a dipole magnetic geometry, one can obtain
W-shaped eclipses similar to those seen for TIC\,141146667.  See in
particular the
\href{http://user.astro.wisc.edu/~townsend/static.php?ref=rrm-movies#Download_Bundles}{movies}
associated with their work; for a $\omega/\omega_{\rm c}$$\approx$0.5,
edge-on system like TIC\,141146667, the eclipses have the correct
approximate shape, and the H$\alpha$ emission can similarly exhibit
double-peaked behavior.

\subsection{Origin of Circumstellar Material and Astrophysical Analogs}

The circumstellar material could originate either from the star or
from an external source.  Plausible external sources include an
undetected old disk, comets, or a close-in exoplanet.  This latter
scenario would make CPVs extrasolar analogs of the Jupiter-Io plasma
torus (e.g.~\citealt{Bagenal1981}), although with a very different
composition.
